---
title: "Exploring the 2015 PHP framework survey by Sitepoint"
author: "Jesus M. Castagnetto"
date: "April 01, 2015"
output: html_document
---

## Background

Sitepoint has published the results of their 2015 PHP Framework popularity survey[^sitepoint]. In that post they show that the survey gives a very large edge to Laravel. The people at Sitepoint were also nice enought to publish their properly anonymized dataset in a github repo[^sitepoint-github] 

[^sitepoint]: http://www.sitepoint.com/best-php-framework-2015-sitepoint-survey-results/

[^sitepoint-github]: https://github.com/sitepoint-editors/php-fw-survey-2015

So I went ahead, and forked their repo and fired up R to give this data a go.

## Exploring the dataset

I read the survey into a data frame, and looked at the overall structure:

```{r warning=FALSE, message=FALSE}
library("dplyr")
library("reshape2")
library("ggplot2")
library("knitr")
library("ROCR")
library("rpart")
library("rpart.plot")
survey <- tbl_df(read.csv("../dump/survey.csv", stringsAsFactors = FALSE))
glimpse(survey)
```

It seems that the first 30 columns related to personal
information and preferences for each respondent (identified by the
`id` field). The next 22 columns comprise a very sparse matrix that
encodes what frameworks (including a write-in option) the respondent
has made contributions.

So I decided to look extract that section of the dataset and do some
quick analysis looking at the frequencies, and how it compared with
the results of framework popularity.

The reason to do that, is to try and get a feeling on the extent to which
each project (framework) has a community that not only benefits from it, but
also works towards improving it by contributing to the code, documentation, etc.

Let's look first at the contributions to the list of selected frameworks:

```{r results='asis'}
# the framework contribution "matrix"
contrib <- survey %>% select(id,laravel:agavi)
ncontrib <- as.data.frame(
  table(rowSums(contrib[, -1], na.rm=TRUE))) %>%
  mutate(pFreq=round(100*Freq/nrow(contrib), 2))
colnames(ncontrib) <- c("Number of frameworks",
                       "Count", "Percent of total")
kable(ncontrib, align=c("c", "c", "c"),
      caption="Distribution of number of frameworks contributed to by respondents")
```

So, about `r 100 - ncontrib[1,3]`% of survey respondents contribute
to at least one framework (without including the "write-ins")

Now, I will convert the data from a "wide" to a "long"
format, using the `reshape2` package, and also mangle the
the write-ins (`other` field) to be able to combine that 
with the other data. 

```{r}
contrib.long <- melt(contrib, id.vars = "id",
                     variable.name = "framework", 
                     value.name = "count",
                     na.rm = TRUE)
head(contrib.long)
summary(contrib.long)
```

It seems that the `count` field is not informative at all, so we can safely
remove it.

```{r}
contrib.long <- contrib.long[,1:2]
```

While mangling the "write-ins", I noticed a large amount of empty
answers, which is reasonable if you consider that the options 
given in the survey contained the most popular frameworks.

```{r}
# the write-ins
contrib.writeins <- survey[,c(1,53)]
nrow(contrib.writeins)
head(contrib.writeins)
# let's remove the empty rows
contrib.writeins <- subset(contrib.writeins, other != "")
nrow(contrib.writeins)
head(contrib.writeins)
```

In the end there are only `r nrow(contrib.writeins)` write-in answers out 
of `r nrow(contrib)` responses. A proportion of about 
`r round(nrow(contrib.writeins)/nrow(contrib) * 100, 2)`% of answers to the survey.

Finally, I combined the two data frames to get all possible 
frameworks to which people are contributing. 
Also, went ahead and removed the string " framework" 
from the framework's name, to help group better, because there were entries
such as "xyz" and "xyz framework".

```{r}
colnames(contrib.writeins) <- c("id", "framework")
contrib.long <- rbind(contrib.long, contrib.writeins)
contrib.long$framework <- gsub(" framework","",
                               contrib.long$framework)
```
### Computing the frequencies of contributions

I used `dplyr` to quickly summarize the results in the long data
frame, calculating the frequency count and percent for each
framework. 

One thing I noticed is that there was a small number of
entries such as: "_i also contribute to php itself_" or
"_took part in customer's development, also fixed codeigniter long tim_", and most of these anomalous "frameworks" were single cases,

Using this summary, I made a bar chart of the top 20 frameworks
to which people contribute, with the package `ggplot2`.

```{r}
contrib.summ <- contrib.long %>% 
  group_by(framework) %>% 
  summarise(freq=n(), pfreq=100*n()/nrow(contrib.long)) %>% 
  arrange(desc(freq))
contrib.summ$framework <- factor(contrib.summ$framework, 
                                levels=rev(contrib.summ$framework),
                                ordered=TRUE)
top20 <- contrib.summ[1:20,]

ggplot(top20, aes(x=framework, y=pfreq)) + 
  geom_bar(stat="identity", fill=rainbow(20), alpha=0.7) +
  geom_text(label=paste0(" ",round(top20$pfreq,1), "%"), hjust=0, size=4.5) +
  xlab("PHP Framework") +
  ylab("Frequency (% of total contributions)") +
  ylim(c(0,18)) +
  ggtitle("Top 20 PHP frameworks by contribution") +
  coord_flip() +
  theme_bw() +
  theme(axis.text=element_text(size=14),
        axis.title=element_text(size=16, face="bold"),
        plot.title=element_text(size=18, face="bold"))
```

## Comparisons with the votes for popular frameworks

Let's now look at the correspondence between frameworks used for work and 
personal projects, and whether the respondent contributes
to at least one framework.

```{r}
personal <- survey[,c(1,5:11,21,30)]
survey$samefw <- survey$fw.work.choice == survey$fw.personal.choice
survey$contributes <- rowSums(contrib[, -1], na.rm=TRUE) > 0
(xt1 <- xtabs( ~ samefw + contributes, survey))
```

Out of every ~7 people who use different frameworks for work and personal projects,
1 contributes to at least one of those frameworks. And, out of ~5 people who
use the same framework at work and personal projects, 1 contributes to at least
one framework.


```{r}
xt2 <- round(100*xt1 / nrow(survey), 1)
```

If we look at this in overall percentages of respondents, we find that:

- `r xt2[1,1]`% use different frameworks for work and personal
  projects, and do not contribute to any framework
- `r xt2[1,2]`% use different frameworks but contribute to at
  least one framework
- `r xt2[2,1]`% use the same framework for work and personal
  projects, but do not contribute to any framework
- `r xt2[2,2]`% use the same framework and contribute to at
  least to one of them

Will there be a difference between contributions for the most popular framework
(according to Sitepoint's 2015 survey): Laravel? How about for the second most
popular: Symfony2?

Again, we use dplyr to summarize this data. First for the use at work, filtering
for the cases where the total number of entries of at least 20 for the framework,
and for entries with an empty name:

```{r}
contrib_fw_work <- survey %>% 
  select(id, fw.work.choice, contributes) %>% 
  group_by(fw.work.choice) %>% 
  summarise(ncontrib=sum(contributes), ntotal=n()) %>% 
  filter(ntotal >= 20 & fw.work.choice !="" ) %>% 
  mutate(pcontrib=ncontrib/ntotal) %>% 
  arrange(desc(pcontrib))
contrib_fw_work %>% filter(fw.work.choice %in% c("Laravel", "Symfony2"))
```

And next for the personal use:

```{r}
contrib_fw_personal <- survey %>% 
  select(id, fw.personal.choice, contributes) %>% 
  group_by(fw.personal.choice) %>% 
  summarise(ncontrib=sum(contributes), ntotal=n()) %>% 
  filter(ntotal >= 20 & fw.personal.choice !="" ) %>% 
  mutate(pcontrib=ncontrib/ntotal) %>% 
  arrange(desc(pcontrib))
contrib_fw_personal %>% filter(fw.personal.choice %in% c("Laravel", "Symfony2"))
```

So Laravel (the most popular framework) has a community engamement (contribution) of
about 15-16%, and Symfony2 something between 23-24%.

If we look at the top 3 frameworks as measured by contributions, the situation
is quite different than the popularity statistics:

```{r results='asis'}
kable(contrib_fw_work[1:3,], digits = 2, 
      caption="Ranking of frameworks used at work by contribution")
```

```{r}
kable(contrib_fw_personal[1:3,], digits = 2, 
      caption="Ranking of frameworks used for personal projects by contribution")
```

It is interesting that these are frameworks that are quite mature and have been
around for quite some time.

## Modeling contribution

The following is not strict modeling, at best it is a preliminary
exploration on trying to figure out if it is possible to 
understand contributions in terms of the variables acquired.

First, we need to calculate the base accuracy, i.e. assigning
to all rows the most frequent value for contribution (`FALSE`).

```{r}
(base_acc <- sum(survey$contributes==FALSE) / nrow(survey))
```

Then, we try a logistic regression (why not?), and let's not
worry about possible colinearity and other hairy issues.

```{r}
survey$age.group <- factor(survey$age.group)
survey$education <- factor(survey$education)
logmodel <- glm(contributes ~ samefw + age.group + years.php +
                years.programming + education + numframeworks,
                data=survey, family="binomial")
summary(logmodel)
logpred <- predict(logmodel, type="response")
(logct <- table(logpred >= 0.5, survey$contributes))
(log_acc <- (logct[1,1] + logct[2,2]) / nrow(survey))
```

The logistic model seem to indicate that the total number of years
of programming experience is not as significative, as, for example,
the number of year programming in PHP, at least to predict 
contributions to PHP frameworks.

As we can see there is a slight improvement in the accuracy using
this naive model, going from `r base_acc` (the base value)
to `r log_acc`, a difference of just
$`r log_acc - base_acc`$

We can now plot and calculate the AUC that we can expect 
with this model.

```{r}
rpred1 <- prediction(logpred, survey$contributes)
perf1 <- performance(rpred1, measure="tpr", x.measure="fpr")
plot(perf1, main="ROC curve for logistic model")
abline(a=0, b=1, lty="dashed", col="blue")
(auc1 <- as.numeric(performance(rpred1, "auc")@y.values))
```

Better than the baseline of 0.5, but not that great. Also,
this model gives us still a great number of false negatives:
people predicted not to contribute, but that do otherwise.

Finally we will try to use a classification tree, and see if 
can get a better model. 


```{r}
treemodel <- rpart(contributes ~ samefw + age.group + years.php +
               years.programming + education + numframeworks,
               data=survey,
               method = "class", cp=0.001)
prp(treemodel)
treepred <- predict(treemodel, type="class")
(treect <- table(treepred, survey$contributes))
(tree_acc <- (treect[1,1] + treect[2,2]) / nrow(survey))
```

The resulting tree is a bit complex, and might be overfitted, and
as we can see, affords a slight improvement in the accuracy, 
going from `r base_acc` (the base value)
to `r log_acc`, a difference of just
$`r log_acc - base_acc`$

Bottomline: I am not convinced that we can model contribution using the
variables collected in this survey. Of course, that was not the goal of
the survey, so it there is no surprise there.
